Exploring Attendance and Performance Trends in Women's Super League

### 2024-12-15 paper/paper.qmd
update paper

---
title: "Exploring Attendance and Performance Trends in Women's Super League"
subtitle: "An analysis of match attendance, key performance factors, and trends in the English Women's Football"
author: 
  - Rayan Awad Alim
thanks: "Code and data are available at: [github.com/RayanAlim/EnglishWomensFootballAnalysis/](https://github.com/RayanAlim/EnglishWomensFootballAnalysis)."
date: December 3, 2024
date-format: long
abstract: "This paper analyzes attendance and performance trends in the Women’s Super League (WSL) using data from the English Women's Football (EWF) Database. We examine the evolution of attendance over time and evaluate how key factors—such as goals scored, goals conceded, and league tier—influence both match-level outcomes and season-level team performance. Using regression modeling, we find that match attendance has grown substantially, correlating with higher probabilities of home-team success, suggesting a measurable 'home advantage.' These insights provide actionable information for enhancing fan engagement strategies and guiding team development, especially as the intrest in womens sports increases globally."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false

@@

Women’s professional football in England (and globally) has experienced substantial growth in recent years, notably reflected in the Women’s Super League (WSL). Increased media coverage, enhanced professionalism, and international success of national teams have all contributed to rising attendance and interest @vanEekeren2022 @Kitching2022. Understanding the determinants of match attendance and performance outcomes is valuable for clubs, team analysts, league organizers, and stakeholders seeking to enhance fan engagement, team performances and competitive standards as demonstrated by @GarciaRodriguez2002, attendance determinants can be significantly influenced by local factors.

@@

This research matters because it provides insights into what drives successful outcomes and robust fan engagement in women’s football. It informs strategic decision-making for clubs aiming to improve both their on-field performances and the experiences they offer to supporters, for organizers to boost engagement, and for analysts interested in understanding sports dynamics. More broadly, these findings are particularly significant as women's sports experience unprecedented growth globally,this study contributes to the growing body of literature that underscores the economic and cultural value of women's football where robust data-driven evidence can shape policy, marketing, and development strategies.

@@

While other datasets, such as FIFA or Opta's proprietary data, exist, they are not as accessible or as granular at the domestic match level. The EWF dataset is chosen for its open availability, rich detail, and focus on the English domestic league structure. This level of granularity (per-match attendance and scores, historical standings) is critical to the research questions here, making the EWF dataset the most suitable for our study.

@@

- Home Team Score Margin (home_team_score_margin) and Away Team Score Margin (away_team_score_margin): Calculated as the difference between goals scored by the respective team and the opponent. These variables provide insight into match competitiveness.

@@

- Attendance (attendance): The number of spectators present at each match. No substantial transformations were needed, although matches without reported attendance were excluded from attendance-specific analyses.The attendance variable captures the number of spectators present at each match, serving as a proxy for fan engagement and the overall popularity of the Women's Super League. Analyzing attendance trends not only provides insights into the growth trajectory of women's football but also highlights key moments that have driven fan interest. Figure @fig-attendance-over-time illustrates the evolution of attendance over time, which shows key shifts and spikes in audience behavior, we can see from @fig-attendance-distribution and @tbl-attendance-summary that attendance is most around the 1000-2000 atendees range. The notable spikes during certain seasons, often corresponding with major international tournaments or high-profile matches. For instance, attendance saw a significant increase following the FIFA Women's World Cup, indicating a spillover effect where global events bolster domestic league interest. This suggests that leveraging international exposure through strategic marketing and scheduling can have a profound impact on league engagement. @tbl-attendance-summary shows the summary of attendance.

@@

The analysis reveals that factors such as the teams playing and the score margin (both home and away) are significant predictors of match attendance. Popular teams and closely contested matches tend to attract more spectators.

@@

The logistic regression results show that higher attendance significantly increases the probability of a home team victory, suggesting the influence of crowd support. Furthermore, the score margins for both teams are critical factors in determining match outcomes.

@@

One of the key findings is that attendance has generally increased over time, with major events and international tournaments serving as pivotal moments that boost interest in women's football. Understanding what drives this interest allows league organizers to align promotional efforts with these catalysts.

@@

## Both offensive and defensive abilities are crucial for a team's performance

Another key observation is the direct impact of goals scored and goals conceded on team standings. Goals scored positively affect match outcomes, as expected, while goals conceded correlate negatively with team success. These results confirm the intuitive idea that both offensive and defensive abilities are crucial for a team's performance. It is not enough to simply score goals-preventing the opponent from scoring also plays an important role.

@@

Moreover, the analysis of attendance reveals that larger crowds tend to coincide with better performances for the home team. This finding highlights the potential advantage that crowd support can provide, supporting the concept of "home-field advantage" in sports.

@@

For organizers, these insights are valuable for shaping promotional strategies that increase attendance, leveraging key calendar events, and enhancing audience experience, thereby boosting overall league engagement and team performance.

@@

Another limitation is the absence of player-specific performance metrics. Incorporating individual-level data, such as player fatigue or injuries, would add depth to our understanding of what drives match outcomes. Future models could incorporate additional predictors to improve predictive power and capture the complex dynamics of match performance.

@@

- **Survey Data:**  
  While this study does not directly use survey data, similar analyses in the literature often rely on surveys to capture fan demographics, motivations, and preferences. Incorporating such surveys into future research could provide richer contextual insights.

@@

### Posterior Predictive Check
A posterior predictive check was conducted to evaluate the extent to which the model captures the patterns in the observed data. In @fig-ppcheckandposteriorvsprior, the observed data is compared against replicated datasets generated from the posterior distribution of the model parameters. This diagnostic provides insights into areas where the model may underfit or overfit the data.

@@

# Residual vs Fitted values
plot(attendance_model, which = 1)
```
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-qq
#| fig-cap: "Diagnostic plots for linear regression assumptions: The Q-Q plot evaluates the normality of residuals, ensuring they follow a standard normal distribution critical for valid inference in linear regression."

>>>>

---
title: "Exploring Attendance and Performance Trends in Women's Super League"
subtitle: "An analysis of match attendance, key performance factors, and trends in the English Women's Football"
author: 
  - Rayan Awad Alim
thanks: "Code and data are available at: [github.com/RayanAlim/EnglishWomensFootballAnalysis/](https://github.com/RayanAlim/EnglishWomensFootballAnalysis)."
date: December 3, 2024
date-format: long
abstract: "This paper analyzes attendance and performance trends in the Women’s Super League (WSL) using data from the English Women's Football (EWF) Database. We examine the evolution of attendance over time and evaluate how key factors—such as goals scored, goals conceded, and league tier—influence both match-level outcomes and season-level team performance. Using regression modeling, we find that match attendance has grown substantially, correlating with higher probabilities of home-team success, suggesting a measurable 'home advantage.' These findings provide actionable information for enhancing fan engagement strategies and guiding team development, especially as the intrest in womens sports increases globally."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false

@@

Women’s professional football in England (and globally) has experienced substantial growth in recent years, notably reflected in the Women’s Super League (WSL). Increased media coverage, enhanced professionalism, and international success of national teams have all contributed to rising attendance and interest @vanEekeren2022 @Kitching2022. Understanding the determinants of match attendance and performance outcomes asissts clubs, team analysts, league organizers, and stakeholders seeking to enhance fan engagement, team performances and competitive standards as demonstrated by @GarciaRodriguez2002, attendance determinants can be significantly influenced by local factors.

@@

This research matters because it provides an outlook into what leads to successful outcomes and robust fan engagement in women’s football. It informs strategic decision-making for clubs aiming to improve both their on-field performances and the experiences they offer to supporters, for organizers to boost engagement, and for analysts interested in understanding sports dynamics. More broadly, these findings are particularly significant as women's sports experience unprecedented growth globally,this study contributes to the growing body of literature that underscores the economic and cultural value of women's football where robust data based evidence can shape policy, marketing, and development strategies.

@@

While other datasets, such as FIFA or Opta's proprietary data, exist, they are not as accessible or as granular at the domestic match level. The EWF dataset is chosen for its open availability, rich detail, and focus on the English domestic league structure. This level of granularity (per-match attendance and scores, historical standings) is needed for the research questions here, making the EWF dataset the most suitable for our study.

@@

- Home Team Score Margin (home_team_score_margin) and Away Team Score Margin (away_team_score_margin): Calculated as the difference between goals scored by the respective team and the opponent. These variables provide an understanding into match competitiveness.

@@

- Attendance (attendance): The number of spectators present at each match. No substantial transformations were needed, although matches without reported attendance were excluded from attendance-specific analyses.The attendance variable captures the number of spectators present at each match, serving as a proxy for fan engagement and the overall popularity of the Women's Super League. Analyzing attendance trends not only provides us information about the growth trajectory of women's football but also highlights key moments that have driven fan interest. Figure @fig-attendance-over-time illustrates the evolution of attendance over time, which shows key shifts and spikes in audience behavior, we can see from @fig-attendance-distribution and @tbl-attendance-summary that attendance is most around the 1000-2000 atendees range. The notable spikes during certain seasons, often corresponding with major international tournaments or high-profile matches. For instance, attendance saw a significant increase following the FIFA Women's World Cup, indicating a spillover effect where global events bolster domestic league interest. This suggests that leveraging international exposure through strategic marketing and scheduling can have a huge impact on league engagement. @tbl-attendance-summary shows the summary of attendance.

@@

The analysis highlights that factors such as the teams playing and the score margin (both home and away) are significant predictors of match attendance. Popular teams and closely contested matches tend to attract more spectators.

@@

The logistic regression results show that higher attendance significantly increases the probability of a home team victory, suggesting the influence of crowd support. Furthermore, the score margins for both teams are important factors in determining match outcomes.

@@

One of the key findings is that attendance has generally increased over time, with major events and international tournaments serving as pivotal moments that boost interest in women's football. Understanding what is the basis of this interest allows league organizers to align promotional efforts with these catalysts.

@@

## Both offensive and defensive abilities are key for a team's performance

Another key observation is the direct impact of goals scored and goals conceded on team standings. Goals scored positively affect match outcomes, as expected, while goals conceded correlate negatively with team success. These results confirm the intuitive idea that both offensive and defensive abilities are key for a team's performance. It is not enough to simply score goals-preventing the opponent from scoring also plays an important role.

@@

Moreover, the analysis of attendance shows us that larger crowds tend to coincide with better performances for the home team. This finding highlights the potential advantage that crowd support can provide, supporting the concept of "home-field advantage" in sports.

@@

For organizers, these findings help in shaping promotional strategies that increase attendance, leveraging key calendar events, and enhancing audience experience, thereby boosting overall league engagement and team performance.

@@

Another limitation is the absence of player-specific performance metrics. Incorporating individual-level data, such as player fatigue or injuries, would add depth to our understanding of what affects match outcomes. Future models could incorporate additional predictors to improve predictive power and capture the complex dynamics of match performance.

@@

- **Survey Data:**  
  While this study does not directly use survey data, similar analyses in the literature often rely on surveys to capture fan demographics, motivations, and preferences. Incorporating such surveys into future research could provide richer contextual understanding

@@

### Posterior Predictive Check
A posterior predictive check was conducted to evaluate the extent to which the model captures the patterns in the observed data. In @fig-ppcheckandposteriorvsprior, the observed data is compared against replicated datasets generated from the posterior distribution of the model parameters. This diagnostic highlights potential areas where the model might underperform or overfit the data.

@@

# Residual vs Fitted values
plot(attendance_model, which = 1)
```
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-qq
#| fig-cap: "Diagnostic plots for linear regression assumptions: The Q-Q plot evaluates the normality of residuals, ensuring they follow a standard normal distribution for valid inference in linear regression."
